---
title: "About the Data"
---

## Introduction

This page shows the workflow to get the cleaned data used in the dashboard starting from the raw data. Happy reading 📚.

### Setup

First, list and load the needed packages.

```{r pkgs, message=FALSE, warning=FALSE}
library(tidyverse)      # to do datascience
library(here)           # to work easily with paths
```

## Read raw data

The raw data are saved in `data/input/20250224_craywatch_raw.txt` as a tab separated text file. It contains GBIF observations of 6 invasive alien crayfishes in Flanders from 2024-06-13, start of the [craywatch](https://craywatch.inbo.be/) project. Derived from [10.15468/dl.pszsrh](https://doi.org/10.15468/dl.pszsrh). The query the GBIF download was based on, is the same query running real-time to show the layer "Nieuwe waarnemingen" (new observations) in the [craywatch map](https://craywatch.inbo.be/kaart/)!

```{r read_data, message=FALSE}
cray_raw_df <- readr::read_tsv(
  here::here("data", "input", "20250224_craywatch_raw.txt"),
  na = "",
  guess_max = 10000
)
```

Preview (10 rows):

```{r preview}
head(cray_raw_df, 10)
```


## Clean data

We remove data without a specific year in column `year` and select only few columns we are interested to:

```{r}
cray_cleaned_df <- cray_raw_df %>%
  filter(!is.na(year)) %>%
  select(gbifID, occurrenceID, species, speciesKey, decimalLatitude,
         decimalLongitude, coordinateUncertaintyInMeters,
         eventDate, year, month, countryCode, stateProvince,
         county, locality, basisOfRecord, individualCount,
         datasetName)
```

Preview (10 rows):

```{r preview_cleaned_data}
head(cray_cleaned_df, 10)
```


## Save cleaned data

We save the cleaned data as `20250224_craywatch_cleaned.txt` in `data/20250224`. This file will be the input file for the dashboard.

```{r}
cray_cleaned_df %>%
  readr::write_tsv(here::here("data",
                              "output",
                              "20250224_craywatch_cleaned.txt")
  )
```
